---
title: "creating the cropped NLCD image"
author: "Quentin D. Read"
date: "5/12/2021"
output: html_document
---

This shows how I used `gdalwarp` to create the cropped NLCD image. This should have been a one liner but `gdalwarp` didn't preserve the attribute table. So I needed to do a workaround and use `terra` to write another image file with attribute table then manually copy that attribute table to the warped raster.

Here use `gdalwarp` with `-tr` (target resolution) to specify that we want 150-m resolution in both directions and `-te` (target extent) to specify the extent (I took it from the dimensions of the `.grd` file used in the old lesson because the script used to create that has been lost to time.)

```{bash, eval = FALSE}
nlcdpath="/nfs/public-data/NLCD/NLCD_2016_Land_Cover_L48_20190424/NLCD_2016_Land_Cover_L48_20190424.img"
gdalwarp -tr 150 150 -te 1394535 1724415 1845135 2101515 -of GTiff -co "COMPRESS=LZW" -overwrite ${nlcdpath} /nfs/qread-data/temp/nlcd_agg.tif
```

So far so good but there is no attribute table. The only way I could figure to make one is read in with terra, use writeRaster to make a dummy image, then copy that image's attribute table to the warped raster.

```{r, eval = FALSE}
library(terra)
nlcd <- rast("/nfs/public-data/NLCD/NLCD_2016_Land_Cover_L48_20190424/NLCD_2016_Land_Cover_L48_20190424.img")

nlcdcrop <- crop(nlcd, ext(c(xmin = 1394535, xmax = 1394635, ymin = 1724415, ymax = 1724515)))
writeRaster(nlcdcrop, '/nfs/qread-data/temp/nlcd_temp.tif')
```

Now copy the attribute table to the new raster.

```{bash, eval = FALSE}
cp /nfs/qread-data/temp/nlcd_temp.tif.aux.xml /nfs/qread-data/temp/nlcd_agg.tif.aux.xml
```

Test.

```{r}
library(stars)
r <- read_stars('/nfs/qread-data/temp/nlcd_agg.tif')
plot(droplevels(r))
```

Dope! Colors and attributes are preserved!


