---
excerpt: Mixing Data Types
---

## Crossing Rasters with Vectors: Prelude

Presently, to mix raster and vectors, we must convert `sf` objects to their
counterpart `Spatial*` objects:

```{r, results = 'hide', handout = 0}
sesync_sp <- as(sesync, "Spatial")
huc_md_sp <- as(huc_md, "Spatial")
counties_md_sp <- as(counties_md, "Spatial")
```

The creation of geospatial tools in R has been a community effort, and not
necessarily a well-organized one. One current stumbling block is that the
[raster](){:.rlib} package, which is tightly integrated with the [sp](){:.rlib}
package, has not caught up to the [sf](){:.rlib} package. The
[stars](https://r-spatial.github.io/stars/) package aims to remedy this problem,
and others, but has not yet released a "version 1.0" (it's at "0.3" as of this
writing).
{:.notes}

===

## Crossing Rasters with Vectors

The `extract` function allows subsetting and aggregation of raster values based
on a vector spatial object.

```{r, fig.keep='last', handout = 0}
plot(nlcd)
plot(sesync_sp, col = 'green',
     pch = 16, cex = 2, add = TRUE)
```

===

When extracting by point locations (i.e. a *SpatialPoints* object), the result
is a vector of values corresponding to each point.

```{r, handout = 0}
sesync_lc <- extract(nlcd, sesync_sp)
```

```{r}
lc_types[sesync_lc + 1]
```

===

When extracting with a polygon, the output is a vector of all raster values for
pixels falling within that polygon.

```{r, handout = 0}
county_nlcd <- extract(nlcd_agg,
    counties_md_sp[1,])
```

```{r}
table(county_nlcd)
```

===

To get a summary of raster values for **each** polygon in a `SpatialPolygons`
object, add an aggregation function to `extract` via the `fun` argument. For
example, `fun = modal` gives the most common land cover type for each polygon in
`huc_md_sp`.

```{r, handout = 0}
modal_lc <- extract(nlcd_agg,
    huc_md_sp, fun = modal)
huc_md_sp$modal_lc <- lc_types[modal_lc + 1]
```

```{r}
head(huc_md_sp)
```
