---
excerpt: Mixing Data Types
---

## Crossing Rasters with Vectors

Raster and vector geospatial data in R require different packages. 
The creation of geospatial tools in R has been a community effort, and not
necessarily a well-organized one. Development is still ongoing on many packages
dealing with raster and vector data.
The [stars](https://r-spatial.github.io/stars/) package has not yet released a "version 1.0" 
(it's at 0.5 as of this writing). Developers are working hard on improving the package but
if you notice any issues or bugs, report them on their [GitHub repo](https://github.com/r-spatial/stars).
{:.notes}

Although not needed in this lesson, you may notice when using some older geospatial packages that it is necessary to
convert a vector object from [sf](){:.rlib} (class beginning with `sfc*`) to [sp](){:.rlib}
(class beginning with `Spatial*`). [sp](){:.rlib} is an older package that was replaced
by [sf](){:.rlib}, and many geospatial packages still work with [sp](){:.rlib} objects.
You can do this by calling `sp_object <- as(sf_object, 'Spatial')`. You may also
need to convert a `stars` object to a `Raster*` class object, which you can do
by calling `stars_object <- as(sf_object, 'Raster')`.
{:.notes}

The `st_extract()` function allows subsetting and aggregation of raster values based
on a vector spatial object. For example we might want to extract the NLCD land cover 
class at SESYNC's location.

```{r, fig.keep='last', handout = 0}
plot(nlcd, reset = FALSE)
plot(sesync, col = 'green',
     pch = 16, cex = 2, add = TRUE)
```

When we use `plot()` to combine raster and vector layers we need to set `reset = FALSE` on 
the first layer we plot so that additional objects can be added to the plot.
{:.notes}

===

Call `st_extract()` on the `nlcd` raster and the `sfc` object containing SESYNC's point location.
When extracting by point locations, the result is another `sfc` with `POINT` geometry and
a column containing the pixel values from the raster at each point.

```{r, handout = 0}
sesync_lc <- st_extract(nlcd, sesync)
```

```{r}
sesync_lc
```

===

Instead of returning a point geometry, you might want to mask a raster by a point geometry instead. 
You can use the indexing operator `[]` to do this. This will set
all pixels in the raster not overlapping the point geometry to `NA`.

```{r}
nlcd[sesync]
```

===

To extract with a polygon, first subset the raster by the polygon geometry. For example
here we subset the NLCD raster by the first row of `counties_md` 
(Baltimore City), resulting in a raster with smaller dimensions that we can plot.

```{r, handout = 0}
baltimore <- nlcd[counties_md[1, ]]
plot(baltimore)
```

===

The expression `nlcd[counties_md[1, ]]` is equivalent to `st_crop(nlcd, counties_md[1, ])`.
The version with `st_crop()` is a bit easier to use with pipes.

===

Here we use a pipe to crop the NLCD raster to the boundaries of Baltimore City,
pull the values as a matrix, and tabulate them.

```{r, handout = 0}
nlcd %>%
  st_crop(counties_md[1, ]) %>%
  pull %>%
  table
```

===

To get a summary of raster values for **each** polygon in an `sfc`
object, use `aggregate()` and specify an aggregation function `FUN`. This
example gives the most common land cover type for each polygon in `huc_md`.

```{r, handout = 0, warning = FALSE}
mymode <- function(x) names(which.max(table(x)))

modal_lc <- aggregate(nlcd_agg, huc_md, FUN = mymode) 
```

The [stars](){:.rlib} package (and base R for that matter) currently have no built-in mode function
so we define a simple one here and pass it to `aggregate()`.
{:.notes}

===

The result is a `stars` object. We can convert it to an `sfc` object containing polygon
geometry.

```{r}
st_as_sf(modal_lc)
```

===

Alternatively, we can extract the values from the `stars` object, add them as a new
column to `huc_md`, and plot:

```{r, handout = 0}
huc_md <- huc_md %>% 
  mutate(modal_lc = modal_lc[[1]])

ggplot(huc_md, aes(fill = modal_lc)) + 
  geom_sf()
```
