---
---

## Interactive Maps

The [mapview](){:.rlib} package is a great way to quickly view vector or 
raster data interactively, using a JavaScript library called leaflet. 
It produces interactive maps (with controls to zoom, pan and toggle
layers) combining local data with base layers from web mapping services.

===

You can pass a vector or raster object to the `mapview()` function, which will create a map
centered over your spatial object over a base tiled map.

By default, it imports tiles from [Carto](https://carto.com/help/building-maps/basemap-list/)
but you can switch to [OpenStreetMap](https://www.openstreetmap.org), 
[ESRI World Imagery](https://www.arcgis.com/home/item.html?id=10df2279f9684e4a9f6a7f08febac2a9), 
or [OpenTopoMap](https://wiki.openstreetmap.org/wiki/OpenTopoMap)
by clicking on the tile icon in the upper left.
{:.notes}

```{r, handout = 0}
library(mapview)
mapview(huc_md)
```

We are using all default arguments for `mapview()`, but there are
many additional options that you can set. You can see a few of them in the
next example.
The map shows up in the "Viewer" tab in RStudio because, like everything with
JavaScript, it relies on your browser to render the image.
{:.notes}

===

Additional spatial datasets (points, lines, polygons, rasters) in your R environment can
be added as map layers using `+`. Mapview will try to make the necessary 
transformation to display your data in [EPSG:3857](https://epsg.io/3857).

```{r, handout = 0}
mapview(nlcd, legend = FALSE, alpha = 0.3, 
        map.types = 'OpenStreetMap') +
    mapview(huc_md, legend = FALSE, alpha.regions = 0.2)
```

If you hover your mouse over one of the polygon features, you can see the polygon ID,
and if you click on it, a pop-up appears with more details. You can zoom and pan the
map with your mouse; restore the zoom to the default or zoom in on one map layer using
one of the buttons at the bottom of the pane.
{:.notes}

Notice also that we passed a few extra arguments to `mapview()` to customize the map. 
Setting `legend = FALSE` gets rid of the large legend panel. Setting
`alpha` for a raster layer and `alpha.regions` for a polygon layer to low values
makes them more transparent so that we can see the underlying map.
Finally, setting `map.types` changes which base map displays by default.
See `?mapview` and `?mapviewOptions` for more options.
{:.notes}

The [mapview](){:.rlib} package uses the R package [leaflet](){:.rlib} behind the 
scenes, which is an R interface to the leaflet JavaScript library. [mapview](){:.rlib}
is great for quick dives into your data, but [leaflet](){:.rlib} provides a lot more
options and customizability if you want to embed maps into websites. For
an in-depth leaflet tutorial, check out the 
[SESYNC lesson on leaflet in R](https://cyberhelp.sesync.org/leaflet-in-R-lesson/).
{:.notes}
