---
---

## Raster Data

Raster data is a matrix or cube with additional spatial metadata (e.g. extent,
resolution, and projection) that allows its values to be mapped onto geographical
space. The [raster](){:.rlib} package provides the eponymous `raster()` function
for reading the many formats of such data.

===

The [National Land Cover Database](http://www.mrlc.gov) is a 30m resolution grid
of cells classified as forest, crops, wetland, developed, etc., across the continental
United States.
The file provided in this lesson is cropped and reduced to a lower resolution in
order to speed processing.

```{r, message = FALSE, handout = 0}
library(raster)
nlcd <- raster("data/nlcd_agg.grd")
```

By default, raster data is *not* loaded into working memory, as you can confirm
by checking the R object size with `object.size(nlcd)`. This means that unlike
most analyses in R, you can actually process raster datasets larger than the RAM
available on your computer; the raster package automatically loads pieces of the
data and computes on each of them in sequence.
{:.notes}

===

The default `print` method for a `raster` object is a summary of metadata contained
in the raster file.

```{r}
nlcd
```

===

The `plot` method interprets the pixel values of the raster matrix according to a
pre-defined color scheme.

```{r, fig.keep = 'last'}
plot(nlcd)
```

===

The `crop()` function trims a raster object to a given spatial "extent" (or
range of x and y values).

Here, we convert the `nlcd` raster's bounding box to a 2x2 matrix of the lower left
and upper right corners, then crop the raster
to the extent of the `huc_md` polygon, then display both layers on the same map.
(We can plot polygon and raster layers together on the same map, just like we can plot
multiple polygon layers---as long as they have the same coordinate reference system!)
Also note that the transformed raster is now loaded in R memory, as indicated by
the size of `nlcd`. We could have also saved the output to disk by specifying an
optional `filename` argument to `crop`; the same is true for many other
functions in the [raster](){:.rlib} package.
{:.notes}

```{r, fig.keep = 'last', handout = 0, warning = FALSE}
extent <- matrix(st_bbox(huc_md), nrow=2)
nlcd <- crop(nlcd, extent)
plot(nlcd)
plot(huc_md, col = NA, add = TRUE)
```

===

A raster is fundamentally a data matrix, and individual pixel values can be
extracted by regular matrix subscripting. For example, the value of
the _bottom_-left corner pixel:

```{r}
nlcd[1, 1]
```

===

The meaning of this number is not immediately clear. For this particular
dataset, the mapping of values to land cover classes is described in the data
attributes:

```{r}
head(nlcd@data@attributes[[1]])
```

Notice the `@` syntax for extracting the attributes from the `nlcd` raster object.
The `@` operator is used to access properties of an object known as "slots." These
slots are nested, so we have to access the slot named `data` from the `RasterLayer`
object, which itself has a slot called `attributes`, which is a list with the attribute
table as its first element(!)
{:.notes}

===

The `Land.Cover.Class` vector gives string names for the land cover type
corresponding to the matrix values. Note that we need to add 1 to the raster
value, since these go from 0 to 255, whereas the indexing in R begins at 1.

```{r, handout = 0}
nlcd_attr <- nlcd@data@attributes 
lc_types <- nlcd_attr[[1]]$Land.Cover.Class
```
```{r}
levels(lc_types)
```

===

### Raster Math

Mathematical functions called on a raster get applied to each pixel. For a
single raster `r`, the function `log(r)` returns a new raster where each pixel's
value is the log of the corresponding pixel in `r`.

Likewise, addition with `r1 + r2` creates a raster where each pixel is the sum of the
values from `r1` and `r2`, and so on. Naturally, spatial attributes of rasters
(e.g. extent, resolution, and projection) must match for functions that operate
pixel-wise on multiple rasters.
{:.notes}

===

Logical operations work too: `r1 > 5` returns a raster with pixel values `TRUE`
or `FALSE` and is often used in combination with the `mask()` function.

```{r, fig.keep = "last", handout = 0}
pasture <- mask(nlcd, nlcd == 81,
    maskvalue = FALSE)
plot(pasture)
```

We use the `mask` function with the logical condition `nlcd == 81` and specify that we want to 
unset pixel values where the mask is false (`maskvalue = FALSE`). This results in a raster 
where all pixels that are not classified as pasture (code `81`) are removed.
{:.notes}

===

To further reduce the resolution of the `nlcd` raster, the `aggregate()`
function combines values in a block of a given size using a given function.

```{r, fig.keep="last", handout = 0}
nlcd_agg <- aggregate(nlcd,
    fact = 25, fun = modal)
nlcd_agg@legend <- nlcd@legend
plot(nlcd_agg)
```

Here, `fact = 25` means that we are aggregating blocks 25 x 25 pixels and `fun =
modal` indicates that the aggregate value is the mode of the original pixels
(averaging would not work since land cover is a categorical variable).
{:.notes}
