---
excerpt: Vector Data
---

## R Packages

The key R packages for this lesson are

- [sf](){:.rlib}
- [raster](){:.rlib}

===

### OSGeo Dependencies

In addition to the common case of depending on other R packages, these two have
dependencies on system libraries.

- [GDAL](https://www.gdal.org) for read/write in geospatial data formats
- [GEOS](https://trac.osgeo.org/geos) for geometry operations
- [PROJ.4](http://proj4.org/) for cartographic projections

System libraries cannot be installed by R's `install.packages()`, but can be
bundled with these packages and for private use by them. Either way, the
necessary libraries are maintained by the good people at the [Open Source
Geospatial Foundation](https://github.com/OSGeo) for free and easy distribution.
{:.notes}

===

### Vector Data

The [US Census
website](http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_county_5m.zip)
distributes county polygons (and much more) that are provided with the handouts.
The [sf](){:.rlib} package reads shapefiles (".shp") and most other vector data:

```{r load_counties, results = 'hide', message = FALSE, handout = 0}
library(sf)

shp <- 'data/cb_2016_us_county_5m'
counties <- st_read(shp, stringsAsFactors = FALSE)
```

===

The `counties` object is a `data.frame` that includes a `sfc`, which stands for
"simple feature column". This special column is usually called "geometry" or
"geom".

```{r sfc}
head(counties)
```

===

### Geometry Types

Like any `data.frame` column, the `geometry` column is comprised of a single
data type. The "MULTIPOLYGON" is just one of several standard geometric data
types.

=== 

| Common Types | Description                                                                                                                  |
|--------------+------------------------------------------------------------------------------------------------------------------------------|
| POINT        | zero-dimensional geometry containing a single point                                                                          |
| LINESTRING   | sequence of points connected by straight, non-self intersecting line pieces; one-dimensional geometry                        |
| POLYGON      | sequence of points in closed, non-intersecting rings; the first denotes the exterior ring, any subsequent rings denote holes |
| MULTI*       | set of * (POINT, LINESTRING, or POLYGON)                                                                                     |

The spatial data types are built upon eachother in a logical way: lines are
built from points, polygons are built from lines, and so on.
{:.notes}

===

We can create any of these spatial objects from coordinates.
Here's a `sfc` object with a single "POINT", corresponding to SESYNC's postition
in WGS84 degrees lat and degrees lon.

```{r st_sfc, handout = 0}
sesync <- st_sfc(st_point(
    c(-76.503394, 38.976546)),
    crs = st_crs(counties))
```

===

### Coordinate Reference Systems

A key feature of a **geo**spatial data type is its associated CRS, stored as an
EPSG ID and an equivalent PROJ.4 string.

```{r}
st_crs(counties)
```

===

### Bounding Box

A bounding box for all featurs in a `sf` data frame is generated by `st_bbox()`.

```{r st_bbox}
st_bbox(counties)
```

===

The bounding box is not a static attribute---it is determined on-the-fly for the
entire table or any subset of features.

```{r st_bbox_md, message = FALSE, handout = 0}
library(dplyr)
counties_md <- filter(
  counties,
  STATEFP == '24')
```

```{r}
st_bbox(counties_md)
```

===

### Grid

A bounding box summarizes the limits, but is not itself a geometry (not a POINT
or POLYGON), even though it has a CRS attribute.

```{r}
st_crs(st_bbox(counties_md))
```

===

A rectangular grid made over a `sf` object is a geometry---by default, a POLYGON.

```{r, handout = 0}
grid_md <- st_make_grid(counties_md,
                        n = 4)
```
```{r}
grid_md
```

===

### Plot Layers

Spatial objects defined by [sf](){:.rlib} are compatible with the `plot`
function. Setting the `plot` parameter `add = TRUE` allows an existing plot to
serve as a layer underneath the new one, so long as the CRS lines up.

```{r plot_counties, fig.keep = "last", handout = 0}
plot(grid_md)
plot(counties_md['ALAND'],
     add = TRUE)
plot(sesync, col = "green",
     pch = 20, add = TRUE)
```

But note that the `plot` function won't prevent you from layering up geometries
with different coordinate systems: you must safegaurd your own plots from this
mistake. The arguments `col` and `pch`, by the way, are graphical parameters
used in base R, see `?par`.
{:.notes}

===

### Spatial Subsetting

An object created with `st_read` is a `data.frame`, which is why the `dplyr`
function `filter` used above on the **non**-geospatial column named "STATEFP"
worked normally. The equivalent of a filtering operation on the "geometry"
column is called a spatial "overlay".

```{r pt_poly_overlay, message = FALSE}
st_within(sesync, counties_md)
```

It can be seen as a type of subsetting based on spatial (rather than numeric or
string) matching. Matching is implemented with functions like `st_within(x, y)`.
The output implies that the 1^st^ (and only) point in `sesync` is within the 5th
element of `counties_md`.
{:.notes}

===

Question
: What was the message issued by the last command all about?

Answer
: {:.fragment} It is a reminder that all geometric calculations are performed as
if the coordinates (in this case longitutde and latitude) are Cartesian x,y
coordinates.

===

The overlay functions in the [sf](){:.rlib} package follow the pattern
`st_predicate(x, y)` and perform the test "x [is] predicate y". Some key
examples are:

| st_intersects | boundary or interior of x intersects boundary or interior of y |
| st_within     | interior and boundary of x do not intersect exterior of y      |
| st_contains   | y is within x                                                  |
| st_overlaps   | interior of x intersects interior of y                         |
| st_equals     | x has the same interior and boundary as y                      |

===

### Coordinate Transforms

For the next part of this lesson, we import a new polygon layer corresponding to
the 1:250k map of US hydrological units (HUC)
[downloaded](http://water.usgs.gov/GIS/dsdl/huc250k_shp.zip) from the United
States Geological Survey.

```{r load_huc, results = 'hide', message = FALSE, handout = 0}
shp <- 'data/huc250k'
huc <- st_read(shp, stringsAsFactors = FALSE)
```

===

Compare the coordinate reference systems of `counties` and `huc`, as given by
their Proj4 strings.

```{r compare_proj_counties}
st_crs(counties_md)$proj4string
```

```{r compare_proj_huc}
st_crs(huc)$proj4string
```

The Census data uses unprojected (longitude, latitude) coordinates, but `huc` is
in an Albers equal-area projection (indicated as "+proj=aea").
{:.notes}

===

The function `st_transform()` converts a `sfc` between coordinate reference
systems, specified with the parameter `crs = x`. A numeric `x` must be a valid
EPSG code; a character `x` is interpretted as a PROJ.4 string.

```{r handout = 0}
prj <- '+proj=aea +lat_1=29.5 +lat_2=45.5 \
    +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 \
    +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 \
    +units=m +no_defs'
```

Proj4 strings contain a reference to the type of projection, this one is another
Albers Equal Area, along with numeric parameters associated with that
projection. An additional important parameter that may differ between two
coordinate systems is the "datum", which indicates the standard by which the
irregular surface of the Earth is approximated by an ellipsoid in the
coordinates themselves.
{:.notes}

===

Use `st_transform()` to assign the two layers to a common projection string
(`prj`). This takes a few moments, as it requires re-calculating coordinates for
every vertex of every polygon in the `sfc`.

```{r handout = 0}
counties_md <- st_transform(
  counties_md,
  crs = prj)
huc <- st_transform(
  huc,
  crs = prj)
sesync <- st_transform(
  sesync,
  crs = prj)
```

===

```{r plot_over, fig.keep = "last", handout = 0}
plot(counties_md$geometry)
plot(huc$geometry,
     border = 'blue', add = TRUE)
plot(sesync, col = 'green',
     pch = 20, add = TRUE)
```

===

### Geometric Operations

The data for a map of watershed boundaries within the state of MD is all here;
in the country-wide `huc` and in the state boundary "surrounding" all of
`counties_md`. To get just the huc in a MD outline:

- remove the internal county boundaries within the state
- clip the hydrological areas to their intersection with the state

===

The first step is a spatial **union** operation: we want the resulting object to
combine the area covered by all the multipolygons in `counties_md`.

```{r st_union, message = FALSE, result = 'hide',  handout = 0}
state_md <- st_union(counties_md)
plot(state_md)
```

To perform a union of all sub-geometries in a single `sfc`, we use the
`st_union()` function with a single argument. The output, `state_md`, is a new
`sfc` that is no longer a column of a data frame. Tabular data can't safely
survive a spatial union and is discarded.
{:.notes}

===

The second step is a spatial **intersection**, since we want to limit the
polygons to areas covered by both `huc` and `state_md`.

```{r handout = 0}
huc_md <- st_intersection(
  huc,
  state_md)
```

===

```{r st_intersect, fig.keep = "last", warning = FALSE, handout = 0}
plot(state_md)
plot(huc_md, border = 'blue',
     col = NA, add = TRUE)
```

The `st_intersection()` function intersects its first argument with the second.
The individual hydrological units are preserved but any part of them (or any
whole polygon) lying outside the `state_md` polygon is cut from the output. The
attribute data remains in the corresponding records of the `data.frame`, but (as
warned) has not been updated. For example, the "AREA" attribute of any clipped
HUC does not reflect the new polygon.
{:.notes}

===

The GEOS library provides many functions dealing with distances and areas. Many
of these are accessible through the [sf](){:.rlib} package, including:

- `st_buffer`: to create a buffer of specific width around a geometry
- `st_distance`: to calculate the shortest distance between geometries
- `st_area`: to calculate the area of polygons

Keep in mind that all these functions use **planar** geometry equations and thus
become less precise over larger distances, where the Earth's curvature is
noticeable. To calculate geodesic distances that account for that curvature,
checkout the [geosphere](){:.rlib} package.
{:.notes}
