---
excerpt: Crossing Types
---

## Crossing Rasters with Vectors: Prelude

Presently, to mix raster and vectors, we must convert needed `sf` objects
to their counterpart `Spatial*` objects:

```{r, results = 'hide', title = '{{ site.handouts[0] }}'}
sesync <- as(sesync, "Spatial")
huc_md <- as(huc_md, "Spatial")
counties_md <- as(counties_md, "Spatial")
```

The creation of geospatial tools in R has been a community effort, and not
necessarilly a well-organized one. One current stumbling block is that the
[raster](){:.rlib} package, which is tightly integrated with the [sp](){:.rlib}
package, has not caught up to the [sf](){:.rlib} package. The
still-under-development [stars](https://r-spatial.github.io/stars/) package aims
to remedy this problem and others.
{:.notes}

===

## Crossing Rasters with Vectors

The `extract` function allows subsetting and aggregation of raster values based
on a vector spatial object.

```{r extract_pt, fig.keep='last', title="{{ site.handouts[0] }}"}
plot(nlcd)
plot(sesync, col = 'green',
     pch = 16, cex = 2, add = TRUE)
```

===

When extracting by point locations (i.e. a *SpatialPoints* object), the result
is a vector of values corresponding to each point.

```{r, title="{{ site.handouts[0] }}"}
sesync_lc <- extract(nlcd, sesync)
```

```{r}
lc_types[sesync_lc + 1]
```

===

When extracting with a polygon, the output is a vector of all raster values for
pixels falling within that polygon.

```{r extract_poly1, title="{{ site.handouts[0] }}"}
county_nlcd <- extract(nlcd_agg,
    counties_md[1,])
```

```{r}
table(county_nlcd)
```

===

To get a summary of raster values for **each** polygon in a `SpatialPolygons`
object, add an aggregation function to `extract` via the `fun` argument. For
example, `fun = modal` gives the most common land cover type for each polygon in
`huc_md`.

```{r extract_poly_agg, title="{{ site.handouts[0] }}"}
modal_lc <- extract(nlcd_agg,
    huc_md, fun = modal)
huc_md$modal_lc <- lc_types[modal_lc + 1]
```

```{r}
head(huc_md)
```
