---
---

## Mixing rasters and vectors

The `extract` function allows subsetting and aggregation of raster values based on a vector spatial object. When extracting by point locations (i.e. a *SpatialPoints* object), the result is a vector of values corresponding to each point.

```{r extract_pt, fig.keep='last', title="{{ site.handouts }}"}
sesync <- spTransform(sesync, proj1)
plot(nlcd)
plot(sesync, col = 'green', pch = 16, cex = 2, add = TRUE)
```

===

```{r}
sesync_lc <- extract(nlcd, sesync)
```

```{r}
lc_types[sesync_lc + 1]
```

===

When extracting with a polygon, the output is a vector of all raster values for pixels falling within that polygon.

```{r extract_poly1, title="{{ site.handouts }}"}
county_nlcd <- extract(nlcd_agg, counties_md[1,])
```

```{r}
table(county_nlcd)
```

===

To get a summary of raster values for **each** polygon in a `SpatialPolygons` object, add an aggregation function to `extract` via the `fun` argument. For example, `fun = modal` gives the most common land cover type for each polygon in `huc_md`.

```{r extract_poly_agg, title="{{ site.handouts }}"}
modal_lc <- extract(nlcd_agg, counties_md, fun = modal)
counties_md$modal_lc <- lc_types[modal_lc + 1]
```

```{r}
head(counties_md)
```
