---
---

## Working with raster data

Raster data is a matrix or cube with additional spatial metadata (e.g. extent, resolution, and projection) that allow its values to be mapped onto geographical space. The **raster** package provides the eponymous `raster()` function for reading the many formats for such data.

===

The [National Land Cover Database](http://www.mrlc.gov/nlcd2011.php) is '.GRD' format data, a lot of it. The file provided in this lesson is cropped and reduced to a lower resolution in order to speed processing.

```{r load_raster, fig.keep="none", title="{{ site.handouts }}"}
library(raster)

nlcd <- raster("data/nlcd_agg.grd")
```

By default, the whole raster is *not* loaded into working memory, as you can confirm by checking the R object size with `object.size(nlcd)`. This means that unlike most analyses in R, you can actually process raster datasets larger than the RAM available on your computer; the raster package automatically loads pieces of the data and computes on each of them in sequence.
{:.aside}

===

```{r}
nlcd
```

===

```{r show_raster, fig.keep="last", title="{{ site.handouts }}"}
plot(nlcd)
```

===

The `crop()` function crops a raster layer to a given spatial *extent* (range of *x* and *y* values). The extent can be extracted from another spatial object with `extent`. Here, we crop the *nlcd* raster to the extent of the *huc_md* polygons, then display both layers on the same map. 

```{r crop_raster, fig.keep="last", title="{{ site.handouts }}"}
nlcd <- crop(nlcd, extent(huc_md))
plot(nlcd)
plot(huc_md, add = TRUE)
```

Note that the transformed raster is now loaded in R memory, as indicated by the size of `nlcd`. We could have also saved the output to disk by specifying an optional `filename` argument to `crop`; the same is true for othe raster transformation functions.
{:.notes}

===

A raster is fundamentally a data matrix, and individual pixel values can be extracted by regular matrix subscripting. For example, this returns the value of the bottom-left corner pixel:

```{r get_raster_values}
nlcd[1, 1]
```

===

The meaning of this number is not immediately clear. For this particular dataset, the mapping of values to land cover classes is described in the data attributes:

```{r raster_attr}
str(nlcd@data@attributes[[1]])
```

===

The `Land.Cover.Class` vector gives string names for the land cover type corresponding to the matrix values. Note that we need to add 1 to the raster value, since these go from 0 to 255, whereas the indexing in R begins at 1.

```{r lc_types}
lc_types <- nlcd@data@attributes[[1]]$Land.Cover.Class
levels(lc_types)
```

===

## Raster math

A mathematical function called on a raster gets applied to each pixel. For a raster `r1`, the function `log(r1)` returns a new raster where each pixel's value is the log of the corresponding pixel in `r1`. The addition of `r1 + r2` creates a raster where each pixel is the sum of the values from `r1` and `r2`, provided their spatial attributes (e.g. extent, resolution, and projection) are the same.

===

Logical operations work too: `r1 > 5` returns a raster with pixel values `TRUE` or `FALSE` and is often used in combination with the `mask()` function. A pasture raster results from unsetting pixel values where the mask (`nlcd == 81`) is false (`maskvalue = FALSE`).

```{r mask, fig.keep="last", title="{{ site.handouts }}"}
pasture <- mask(nlcd, nlcd == 81, maskvalue = FALSE)
plot(pasture)
```

===

To further reduce the resolution of the `nlcd` raster, the `aggregate()` function combines values in a block of a given size using a given function.

```{r agg_raster, fig.keep="last", title="{{ site.handouts }}"}
nlcd_agg <- aggregate(nlcd, fact = 25, fun = modal)
nlcd_agg@legend <- nlcd@legend
plot(nlcd_agg)
```

Here, `fact = 25` means that we are aggregating blocks 25 x 25 pixels and `fun = modal` indicates that the aggregate value is the mode of the original pixels (averaging would not work since land cover is a categorical variable).
{:.notes}

===

## Exercise 3

The function `cellStats` aggregates accross an entire raster. Use it to figure out the  proportion of `nlcd` pixels that are covered by deciduous forest (value = 41).

[View solution](#solution-3)
